Tropical mathematics for the Lambda-calculus 1

### 2023-07-28 CSL24/section2bis.tex
Commit pre-cena

\begin{remark}
 There is a clear and well-known relation between tropical polynomial/power series and usual polynomials/power series:
fix a (formal) power series $f=\sum_{i\in I\subseteq\N} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$;
fix a \emph{valuation} $\mathrm{val}:Q\to \Lawv$.
Remember that a valuation by definition always sends $1\in Q$ to $0$.
Then one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.
The constantly equal to $0$ valuation is called the \emph{null valuation}.
A common valuation on $Q=[0,1]$ is $\mathrm{val}(a)=-\log a$.
\end{remark}

>>>>

\begin{remark}\label{rmk:val_trop}
 There is a clear and well-known relation between tropical polynomial/power series and usual polynomials/power series. 
Given $Q,L$ semirings with units and zeros, $L$ commutative idempotent, if $L$ is totally ordered by the partial order: $\alpha \preceq \beta$ iff $\alpha +  \beta = \beta$, then one defines a \emph{valuation} \cite{Izhakian2015} to be a map $\mathrm{val}:Q\to L$ s.t.\ $\mathrm{val}(0)=0$, $\mathrm{val}(1)=1$, $\mathrm{val}(ab)=\mathrm{val}(a)\mathrm{val}(b)$, $\mathrm{val}(a+b)\preceq \max\set{\mathrm{val}(a),\mathrm{val}(b)}$.
The valuation which is constantly equal to $1\in L$ on all the invertibles of $Q$ is called the \emph{trivial valuation} $\mathrm{val}_1$.

Now, for a (formal) power series $f=\sum_{i\in I\subseteq\N} a_i x^i$ (polynomials being the case for $I$ finite) with coefficitents $a_i$ in a semiring $Q$, and for a fixed valuation $\mathrm{val}:Q\to \Lawv$ (notice that in $\Lawv$ we have that $\preceq$ is $\geq$ and $\max^{\preceq}=\min$), one defines the \emph{tropicalisation} $\trop^{\mathrm{val}}f:\Lawv \to \Lawv$ of $f$ as the tropical polynomial/power series function $\trop^{\mathrm{val}}f(\alpha):=\inf_{i\in I} \set{\mathrm{val}(a_i)+i\alpha}$.

Remark that tropicalising w.r.t.\ the trivial valuation gives $\trop^{\mathrm{val}_1} \sum_{i\in I} a_i x^i = \inf_{i\in I} \set{i\alpha} = \trop^{\mathrm{val}} \sum_{i\in I\subseteq\N} x^i$ for all valuation $\mathrm{val}$, because in $\Lawv$ the unit is $0$. %and the zero is $\infty$.
%Remember that a_i \neq 0 by def of formal power series
That is, it coincides to consider the same power series but with coefficients in $Q$ quotiented for the idempotency of $+$.

\end{remark}

### 2023-07-28 CSL24/section4.tex
Commit pre-cena

The results just presented translate into the following facts about the interpretation of higher-order programs.
Remark that the metric spaces $(\Lawv^{\mathlarger{+}_{i=1}^n X_i},d_{\infty})$ and $(\prod_{i=1}^n \Lawv^{X_i},\max_{i=1}^n d^{X_i}_\infty)$ are trivially isometric, so we identify them.

>>>>

\subsection{The tropical interpretation of high-order programs}\label{subsec:tro_int}

Rcall that the Taylor expansion $\Te{MN}$ of $MN$ is the set $\set{\Der^{n}[t,u_1,\dots,u_n]0 \mid n\in\N,t\in\Te{M},u_i\in\Te{N}}$.
%One defines its interpretation as $\model{\Te{MN}}:=\inf_{n\in\N,t\in\Te{M},u_i\in\Te{N}} \model{\Der^{n}[t,u_1,\dots,u_n]0}$.
$\LREL_!$ is also well-behaved w.r.t.\ the Taylor expansion, as expressed by the following two results.
First, it can be patientely checked that in $(\LREL_!,D)$ all morphisms can be Taylor expanded  (see \cite[Definition 4.22]{Manzo2012}):

\begin{theorem}\label{thm:modelsTaylor}
 For all $t\in\HOM{\LREL_!}{Z}{X\multimap Y}$, $s\in\HOM{\LREL_!}{Z}{X}$ we have:%, the evaluation of $t$ over $s$ yields 
 \begin{align}\label{eq:taylorcat}
  \RM{ev}\circ_!\langle t,s\rangle =
  \inf\limits_{n\in\N}
  \set{((\dots((\Lambda^- t)\star s)\star \dots)\star s)\circ_! \langle \RM{id},\infty \rangle}.
 \end{align} 
\end{theorem}
%It is worth discussing the formula above a bit more. 
Here,
%:\HOM{\LREL}{!Z}{X\multimap Y}\to \HOM{\LREL}{!(Z+X)}{Y}$
%$\star:\HOM{\LREL}{!(Z+X)}{Y}\times\HOM{\LREL}{!Z}{X}\to \HOM{\LREL}{!(Z+X)}{Y}$ is defined as 
$u\star s= (Du)\circ_{!} \langle \langle  \infty, s\circ_{!} \pi_{1}\rangle,\mathrm{id}\rangle$ corresponds to the application of the derivative of $u$ on $s$, and $\Lambda^-$ is the uncurry operator.
Hence the right-hand term in \eqref{eq:taylorcat} corresponds to the $\inf$ of the $n$-th derivative of $\Lambda^{-}t$ applied to ``$n$ copies'' of $s$,  i.e.~it coincides with the tropical %interpretation of the
version of the usual Taylor expansion.
%Moreover, since $\LREL_!$ has countable sums (all $\inf$'s converge), and thanks to equation \eqref{eq:taylorcat}, an immediate adaptation of the proof of [Theorem 4.23, \cite{Manzo2012}] entails that the interpretation of the $\STDLC$-Taylor expansion of a $\STLC$-term $M$ given in \eqref{eq:taylor}, converges to the interpretation of $M$.
Second, since $\LREL_!$ has countable sums (all countable $\inf$s converge), an immediate adaptation of the proof of \cite[Theorem 4.23]{Manzo2012} shows:

\begin{corollary}\label{cor:T(M)=M}
 For $\Gamma\vdash_{\STLC} M:A$, we have %$\model{\Gamma\vdash_{\STDLC} \Te M:A}:=
$\model{\Gamma\vdash_{\STLC} M:A}=\inf_{t\in\Te{M}} \model{\Gamma\vdash_{\STDLC} t:A}$. %the interpretation of the Taylor expansion of a $\STLC$-term $M$, given in \eqref{eq:taylor}, converges to the one of $M$.
\end{corollary}

%More formally, the differential operator $\Der[-]$ transforms a function $M:A\to B$ into a function $\Der[M]: A\to (A\to B)$ which is linear in its first argument. 
%Since $M$ may rather ask for several copies of $N$, this requires a form of non-determinism: 
%For example, if $M$ is the term $\lambda fx.f(fx)$ considered before, $\Der[M]$ takes a first input $N$ and passes it linearly to $M$. Notice that there are two ways of doing so, corresponding to the two bound occurrences of $f$ in $M$: either by applying $N$ to $fx$, or by 
%applying $f$ linearly to $Nx$ (indeed, if $f$ were applied in an unrestricted way, it might duplicate $Nx$, so that $N$ would not be used linearly). This justifies the equation below, in which $\Der[M]$ is identified with the non-deterministic sum of the two possible linear choices:
%\begin{align}
%\Der\left[\lambda f x.f(fx)\right]\cdot N = 
%\lambda fx. N(fx) + \left(\Der[f]\cdot (Nx)\right)(fx)
%\end{align}
%More generally, one can define a notion of $k$-bounded application $\Der^{(k)}[M]\cdot N^{k}$, where $\Der^{(0)}[M]\cdot N^{0}= M$ and $\Der^{(k+1)}[M]\cdot N^{k+1}= \Der[ \Der^{(k)}[M]\cdot N^{k}]\cdot N$, corresponding to passing $N$ to $M$ exactly $k$ times.
%
%
%The name ``differential'' for the operator $\Der[-]$ is justified by the fact that it satisfies many properties of the usual differential operator of analysis $\Der[f]:= \lambda xy. \frac{\mathsf df(y)}{\mathsf dy}\cdot x$. Notably, it is additive in its first variable (i.e.~it commutes with the non-deterministic sum operator), and satisfies the chain rule.
%Most famously, the differential operator can be used to define a Taylor formula for $\lambda$-terms, which decomposes an unrestricted application into a formal non-deterministic sum of bounded applications:

%
%More generally, the relational semantics interprets unbounded programs as \emph{analytic functions}, that is, as functions admitting a representation as power series. For instance, observing that an analytic map $f: \BB R\to \BB R$, where $f(x)=\sum_{n}\widehat f_{n}\cdot x^{n}$ is uniquely determined by the sequence $\widehat f_{n}$, the program $M_{\infty}:=\lambda fx.fx: (\BB R\To \BB R)\To (\BB R\To \BB R)$ is represented by the power series below:
%\begin{align}
%F_{\infty}(f,x)= \sum_{n=0}^{\infty} \widehat f_{n} x^{n}
%\end{align}
%By restricting ourselves to bounded applications, the terms in the power series become finite, that is, the interpretation becomes a \emph{polynomial}: for instance, the program $M_{2}:=\lambda fx. \sum_{i=0}^{2}\Der^{(i)}[f]\cdot x^{i}$, corresponding to passing $x$ \emph{at most twice} to $f$, is represented by the polynomial
%\begin{align}
%F_{\leq 2}(f,x)=\widehat f_{2} x^{2}  + \widehat f_{1}x +  \widehat f_{0} 
%\end{align}
% In this framework the differential operator is naturally represented by formal differentiation of polynomials, where, as one would expect, 
% $\Der[\sum_{n}a_{n}x^{n}]=\sum_{n}\Der[a_{n}x^{n}]$ and $\Der[a_{0}x^{0}]=0$ and $\Der[a_{n+1}x^{n+1}]= (n+1)a_{n+1}x^{n}$, so that power series can be Taylor expanded.

The results presented in all this paper translate now into the following facts about the interpretation of higher-order programs.
Remark that the metric spaces $(\Lawv^{\mathlarger{+}_{i=1}^n X_i},d_{\infty})$ and $(\prod_{i=1}^n \Lawv^{X_i},\max_{i=1}^n d^{X_i}_\infty)$ are trivially isometric, so we identify them.

### 2023-07-28 CSL24/section5.tex
Commit pre-cena

Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega_0}, Q_{\omega_0}$, which is more easily solved by 
passing to the tropical monomial/polynomial functions $\trop^{\mathrm{val}} P_{\omega_0},\trop^{\mathrm{val}} Q_{\omega_0}$. 
For 1), by definition of $\RM{arg max}$ and because $-\log$ is stricly decreasing, we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)\in
%\begin{equation}
%  \begin{array}{ccccc}
   \RM{arg max}_{(x,y)} P_{\omega_0} (x,y)
   %& 
   = %&
   \RM{arg min}_{(x,y)}\set{-\log P_{\omega_0} (x,y)}
   %&
   =%&
   \RM{arg min}_{(x,y)}\set{(\trop P_{\omega_0}) (-\log x,-\log y)} \label{eq:argmax}$
where this holds for any valuation.
Remark that $(\trop P_{\omega})( -\log x, -\log y)$ is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow \false_{\omega}$'', se we see that the tropicalisation allows to compute such quantities.
%  \end{array}
%\end{equation}
For 2), %the $\omega_l$ is s.t.\ $\max\limits_{\omega\in\set{ll,rll,rrr}} \, P_\omega(x,y) = P_{\omega_l}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max_{\omega\in\set{lr,rrl,rlr}} \, P_\omega(p,q) = P_{\omega_0}(p,q)$, i.e.\ $\min_{\omega\in\set{lr,rrl,rlr}} \, -\log P_\omega(p,q) = -\log P_{\omega_0}(p,q)$.
Remembering that $\trop^{\mathrm{val}} Q_{\false}(p,q)=\min\set{p+q, \mathrm{val}(2)+p+2q}$, we see that our minimization problem is equivalent to the equality $(\trop^0 Q_{\false})( -\log p, -\log q) = (\trop^0 P_{\omega_0})( -\log x, -\log y)\label{eq:max}$
and this holds only for the null valuation.
Remark that, in both cases, passing through $\trop^0 $ %P_\omega, \trop Q_\omega$ 
makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2lr5}), and this basically corresponds to study negative log-probabilities. %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

@@

\begin{remark}\label{rmk:tropof01Rel}
This toy first-order language can be interpreted in the already mentioned $\overline{R_{\geq 0}}\mathrm{Rel}$ and in $\LREL$.
We do not give details now since, from the following section, we will introduce the interpretation of interesting high-order calculi, including a probabilistic one containing the one of this section.
But it is important to mention already at this point that the probabilities are already captured by the $[0,\infty]\mathrm{Rel}$ model: the $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}\in[0,\infty]^{\set{0,1}}$ of our running example $M$ is $\model M_0=Q_{\false}(p,1-p)$, $\model M_1=Q_{\true}(p,1-p)$.
Therefore, this optimisation problems are already expressible by taking $\trop^0\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$. 
Now, the model $\LREL$ is precisely null-valuation tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$, i.e.\ $\model{M}^{\LREL}=\trop^0\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$.
This corresponds to quotienting the polynomial interpreting $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$ w.r.t.\ idempotent sum, as the null valuation eliminates all the coefficients (different from $1$).
A precise study of the ``tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$'' is left for future investigations, as it is related with power series arising from more sophisticated calculi with paramethers (in the style of \cite{} {\color{red}Ehrhard!}).
We therefore have that $\model{M}^{\LREL}_0(-\log p,-\log (1-p)$ is the negative log-probability of \emph{any} of the (equiprobable) most likely paths of reduction to normal form.
We take these observations as a motivation for considering such model, and the point of this work is to dig into such model, concentrating for this first paper on the relations between the metric and differential tools that we will consider.
% (in $p$ and $q$, not after the substitution $q:=r-p$) 
% are extracted from it, \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
\end{remark}

>>>>

Answering 1) and 2) amounts at solving a maximization problem related to $P_{\omega_0}, Q_{\omega_0}$, which is more easily solved by 
passing to the tropical monomial/polynomial functions $\trop P_{\omega_0},\trop^{\mathrm{val}} Q_{\omega_0}$, for any fixed valuation.
For 1), by definition of $\RM{arg max}$ and because $-\log$ is stricly decreasing, we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and $(p,q)\in
%\begin{equation}
%  \begin{array}{ccccc}
   \RM{arg max}_{(x,y)} P_{\omega_0} (x,y)
   %& 
   = %&
   \RM{arg min}_{(x,y)}\set{-\log P_{\omega_0} (x,y)}
   %&
   =%&
   \RM{arg min}_{(x,y)}\set{(\trop P_{\omega_0}) (-\log x,-\log y)} \label{eq:argmax}$
where this holds for any valuation (because monomials do not carry any coefficient).
Remark that $(\trop P_{\omega})( -\log x, -\log y)$ is precisely the \emph{negative log-probability} of the event ``$M\twoheadrightarrow \false_{\omega}$'', so we see that the tropicalisation allows to compute such quantities.
%  \end{array}
%\end{equation}
For 2), %the $\omega_l$ is s.t.\ $\max\limits_{\omega\in\set{ll,rll,rrr}} \, P_\omega(x,y) = P_{\omega_l}(x,y)$. So
we are looking for $p,q\in[0,1]$ s.t.\ $q=1-p$ and
$\max_{\omega\in\set{lr,rrl,rlr}} \, P_\omega(p,q) = P_{\omega_0}(p,q)$, i.e.\ $\min_{\omega\in\set{lr,rrl,rlr}} \, -\log P_\omega(p,q) = -\log P_{\omega_0}(p,q)$.
Remembering that $\trop^{\mathrm{val}} Q_{\false}(p,q)=\min\set{p+q, \mathrm{val}(2)+p+2q}$, we see that our minimization problem is equivalent to asking $(\trop^{\mathrm{val}_1} Q_{\false})( -\log p, -\log q) = (\trop^{\mathrm{val}_1} P_{\omega_0})( -\log x, -\log y)\label{eq:max}$,
and this time this holds, in general, only for the trivial valuation.
Remark that, in both cases, passing through $\trop^0 $ %P_\omega, \trop Q_\omega$ 
makes the problem easier, as this amounts to study tropical polynomials (for instance computing tropical roots can be done in linear time \cite{Noferini2lr5}), and this essentially corresponds to study negative log-probabilities. %the tropicalisation operator $\trop{}$ as well as the \emph{negative $\log$-probabilities} appear.

@@

\begin{remark}\label{rmk:tropof01Rel}
Our toy first-order language can be interpreted both in the already mentioned $\overline{R_{\geq 0}}\mathrm{Rel}$ and in $\LREL$.
We do not give details now since in the rest of the paper we will interpret interesting \emph{high-order} calculi, including a probabilistic one containing our toy language.
But it is important to mention already at this point that the probabilities we just discussed above, are already captured by the $[0,\infty]\mathrm{Rel}$ model: the $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}\in[0,\infty]^{\set{0,1}}$ of our running example $M$ is $\model M_0=Q_{\false}(p,1-p)$, $\model M_1=Q_{\true}(p,1-p)$.
Therefore, the optimisation problems concerning the negative log-probabilities and likelihoods are already expressible by taking $\trop^{\mathrm{val}_1}\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$. 
Now, the model $\LREL$ is precisely trivial-valuation tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$, i.e.\ $\model{M}^{\LREL}=\trop^{\mathrm{val}_1}\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$.
As we remarked in \autoref{rmk:val_trop}, this corresponds to quotienting the polynomial interpreting $\model{M}^{\overline{R_{\geq 0}}\mathrm{Rel}}$ w.r.t.\ idempotent sum.
A precise study of the ``tropicalisation of $\overline{R_{\geq 0}}\mathrm{Rel}$'' is left for future investigations, as it is related with power series arising from more sophisticated calculi with paramethers (in the style of \cite{} {\color{red}Ehrhard!}).
For now, just have in mind that $\model{M}^{\LREL}_0(-\log p,-\log (1-p))$ gives the negative log-probability of \emph{any} of the (equiprobable) \emph{most likely} reduction paths to normal form, so $\LREL$ directly expresses such optimisations problems (no need to pass through a tropicalisation).
We take these observations as incouraging, and the point of this work is to dig into such model, concentrating for this first paper on the relations between the metric and differential properties that coexist in it.
% (in $p$ and $q$, not after the substitution $q:=r-p$) 
% are extracted from it, \ref{eq:argmax}, \ref{eq:max} of $\STLC_\oplus$-programs. 
\end{remark}
